Learning action models with minimal observability

### 2018-07-24 doc/aij18/aij18.tex
Add files via upload
Introduction done

\newtheorem{mytheorem}{Theorem}
\newtheorem{mylemma}[mytheorem]{Lemma}
\newtheorem{mydefinition}[mytheorem]{Definition}
\newtheorem{myconstruction}{Construction}

@@

\begin{abstract}
  This paper presents a novel approach for learning \strips\ action models from observations of plan executions that compiles this learning task into classical planning. The compilation approach is flexible to various amount and kind of available input knowledge; learning examples can range from plans (with their corresponding initial state) to sequences of state observations or even just a set of initial and final states (where no intermediate action or state is known). The compilation accepts also partially specified action models and can be used to validate whether the observation of a plan execution follows a given \strips\ action model, even if this model is not fully specified. On the other hand, the compilation is extensible to assess how well a given \strips\ action model matches the observation of a plan execution. This extension allows us to evaluate the quality of the learned models with respect to the actual models but also, with respect to a {\em test set} of observations of plan executions. The performance of our compilation approach is evaluated learning action models, for a wide range of classical planning domains from the International Planning Competition (IPC), and following these two evaluation approaches.
\end{abstract}

\begin{keyword}
Classical planning\sep Learning action models\sep Generalized planning
%% keywords here, in the form: keyword \sep keyword

>>>>

\newtheorem{mytheorem}{Theorem}
\newtheorem{mylemma}[mytheorem]{Lemma}
\newtheorem{mydefinition}[mytheorem]{Definition}
\newtheorem{myproposition}[mytheorem]{Proposition}
\newtheorem{myconstruction}{Construction}

@@

\begin{abstract}
  \textcolor[rgb]{1.00,0.00,0.00}{This paper presents a novel approach for learning \strips\ action models from observations of plan executions that compiles this learning task into classical planning. The compilation approach is flexible to various amount and kind of available input knowledge; learning examples can range from plans (with their corresponding initial state) to sequences of state observations or even just a set of initial and final states (where no intermediate action or state is known). The compilation accepts also partially specified action models and can be used to validate whether the observation of a plan execution follows a given \strips\ action model, even if this model is not fully specified. On the other hand, the compilation is extensible to assess how well a given \strips\ action model matches the observation of a plan execution. This extension allows us to evaluate the quality of the learned models with respect to the actual models but also, with respect to a {\em test set} of observations of plan executions. The performance of our compilation approach is evaluated learning action models, for a wide range of classical planning domains from the International Planning Competition (IPC), and following these two evaluation approaches.}
\end{abstract}

\begin{keyword}
Action model learning\sep AI planning\sep Machine Learning
%% keywords here, in the form: keyword \sep keyword

### 2018-07-24 doc/aij18/evaluation.tex
Introduction done

The \ARMS system showed that a semantic evaluation can be done via validation of a set of plan traces with the learned model~\cite{yang2007learning}. The underlying idea is that an error indication of the learned action models is obtained by counting the number of preconditions that are not satisfied during the execution of the plan trace with the learned models. This approach can be understood as modifying the plan trace (by adding the necessary preconditions to the intermediate states) so as to allow the execution of the observed actions using the learned models. In other words, modifying the plan trace to fit the model. Inspired by this approach, we present here an alternative evaluation that, instead, determines the modifications required by a learned model to explain the given plan traces.

>>>>

The \ARMS system showed that a semantic evaluation can be done via validation of a set of plan traces with the learned model~\cite{yang2007learning}. The underlying idea is that an error indication of the learned action models is obtained by counting the number of preconditions that are not satisfied during the execution of the plan trace with the learned models, similarly to the functionality provided by the automatic validation tool VAL~\cite{howey2004val} used in the IPCs. This approach can be understood as modifying the plan trace (by adding the necessary preconditions to the intermediate states) so as to allow the execution of the observed actions using the learned models. In other words, modifying the plan trace to fit the model. Inspired by this approach, we present here an alternative evaluation that, instead, determines the modifications required by a learned model to explain the given plan traces.

### 2018-07-24 doc/aij18/introduction.tex
Introduction done

Besides {\em plan synthesis}~\cite{ghallab2004automated}, planning action models are also useful for {\em plan/goal recognition}~\cite{ramirez2012plan}. At both planning tasks, automated planners reason about action models that correctly and completely capture the possible world transitions~\cite{geffner:book:2013}. Unfortunately, modeling planning actions is complex, even for planning experts, and this knowledge acquisition task is a bottleneck that limits the potential of AI planning~\cite{kambhampati:modellite:AAAI2007}.

Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models, the vanilla action model for automated planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The {\em input} to ML algorithms (the {\em learning/training} data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each observed plan possibly has a different number of steps and involves a different number of objects).
\item The {\em output} of ML algorithms usually is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of {\em preconditions}, {\em negative} and {\em positive effects} that define the possible state transitions.
\end{itemize}

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016generalized,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. A solution to the classical planning task that results from our compilation is a sequence of actions that determines the preconditions and effects of a \strips\ model such that this model satisfies the plan executions given as input.

The compilation approach is appealing by itself because it leverages off-the-shelf planners and because its practicality allow us to report learning results over a wide range of domains from the {\em International Planning Competition} (IPC). Moreover, it opens up a way towards \emph{bootstrapping} planning action models, enabling a planner to gradually learn/update its action model. Apart from these, our compilation exhibits the following contributions:
\begin{enumerate}
\item {\bf\em Input flexibility}. Our classical planning compilation is flexible to various amount and kind of available input knowledge. The action model to learn can be partially specified and the learning examples can range from a set of plans (with their corresponding initial state) or state observations, to just a set of initial and final states where no intermediate action or state is observed. % Also accepts background knowledge

\item {\bf\em Model validation}. The compilation poses a novel framework to assess the validation of a STRIPS model with respect to plan executions. This validation capacity goes beyond the functionality of VAL~\cite{howey2004val} since it does require neither a full action model nor a fully observed plan to determine validation.

\item {\bf\em Model evaluation}. The compilation can assess how well a given \strips\ action model matches a plan execution, which allows us to assess learning performance without knowing the actual action model. The idea is to assess the amount of edition that is required by the input action model to induce the given observations of plan executions. 
\end{enumerate}

A first description of the compilation previously appeared in our previous conference paper~\cite{aineto2018learning}. Compared to that paper, this work includes the following novel material:
\begin{itemize}
\item A unified formulation for learning and evaluating \strips\ action models from observed executions of plans. Further, these executions can only comprise state observations.
\item A redefinition of the ML metrics {\em precision} and {\em recall} to evaluate \strips\ action models with respect to observations of plan executions.
%\item Leveraging diverse forms of {\em background knowledge} for learning and evaluating \strips\ action models.
\item A complete empirical evaluation of the compilation approach for learning of \strips\ action models. Our evaluation analyses how input knowledge affects to the performance of the compilation approach when learning and evaluating \strips\ action models. 
\end{itemize}

Section~\ref{sec:background} introduces classical planning and reviews related work on learning planning action models. Section~\ref{sec:motivation} motivates our compilation approach for learning of \strips\ action models from observations of plan executions. Section~\ref{sec:learning} introduces the \strips\ action model (the target of our learning and evaluation tasks) and formalizes the learning of \strips\ action models with regard to different amount and kind of available input knowledge. Sections~\ref{sec:Section5} and ~\ref{sec:Section6} describe our compilation approach for addressing the formalized learning tasks its extension to evaluate and recognize \strips\ action models. Section~\ref{sec:Section7} reports the data collected in a two-fold empirical evaluation of our learning approach: First, the learned \strips\ action models are tested with observations of plan executions and second, the learned models are compared to the actual models. Finally, Section ~\ref{sec:Section8} discusses the strengths and weaknesses of the compilation approach and proposes several opportunities for future research.

>>>>

There is common agreement in the planning community that the unavailability of a complete domain model is a bottleneck in the applicability of planning technology to many real-world domains~\cite{kambhampati:modellite:AAAI2007}. Motivated by the difficulty and cost of crafting action models, research in action-model learning has seen huge advances since the emergence of the pioneer learning system ARMS~\cite{yang2007learning}. Ever since we have seen models able to learn action models with quantifiers~\cite{AmirC08,ZhuoYHL10}, from noisy actions or noisy states~\cite{zhuo2013action,MouraoZPS12}, from null state information~\cite{cresswell2013acquiring}, from incomplete domain models~\cite{ZhuoNK13,ZhuoK17} and many more.

The primary underlying motivation for acquiring planning action models is to be able to solve model-based planning tasks afterwards. A learning system receives as an input observations of the agent's plan execution and outputs an abstract version of the capability model that reflects the physics of the real-world domain. Recently, model-based representations are also becoming popular in explainable AI planning as they form a common basis for communicating with users and facilitate the generation of transparent and explainable decisions~\cite{FoxLM17} as well as explanations in terms of the differences with a human mental model~\cite{ChakrabortiSK18}.

Two different types of data sources are generally identified in the generation of explanations~\cite{Sheh17,Roberts18}: (1) data that embody enough information to fully trace the steps of the decision-making process of the agent and so accurate and detailed explanations can be generated; (2) data that do not accurately represent the decision-making process because the observed external behaviour of the agent responds to a black-box model. A similar distinction can be established with the type of observations that algorithms use to learn planning action models. Up to date, all existing learning approaches assume that a given input plan trace contains a fully observed sequence of the executed actions. This seriously restricts the applicability of the learning approach to contexts in which the behaviour of the agent is fully observable and a human annotator correctly labels the observed actions.

In this paper we claim that a planning model is learnable even though an accurate representation of the agent's behaviour is not available. Particularly, we present a novel learning algorithm, called \FAMA, capable of inferring the preconditions and effects of \strips\ action models, the vanilla action model for automated planning~\cite{fikes1971strips}, under a minimum input knowledge. While current learning systems accept input plan traces to a greater or lesser degree of observability in the states traversed by the sequence of observed actions, \FAMA extends the degree of observability to the action sequence. This way, we can define a percentage of observability of the actions up to a null observability. The minimal observability case acceptable by \FAMA is when the algorithm is only fed with the input and output of a plan trace, namely the initial and final state of a planning task. Like many Machine Learning (ML) techniques, \FAMA is able to operate with only correct input/ouput pairs and an unknown or a partially known model of the agent. Unlike ML algorithms, \FAMA requires a symbolic structured representation of the input knowledge. In this sense, recent investigations tackle the problem of learning symbolic representations from low-level sensing information and unstructured data~\cite{KonidarisKL18,AsaiF18}.

\FAMA is a solving schema, based on AI planning technology, that automatically compiles the task of learning \strips\ actions into a planning task which is then solved with a planner~\cite{aineto2018learning}. \FAMA is thereby a model-based approach that automatically builds its own planning model by logical inference from the input observations. A solution to this planning task is a sequence of actions that determines the preconditions and effects of the \strips\ action models. The construct of the planning model starts out with the input observations of the agent's behaviour that, as commented above, may comprise none of the actions executed by the agent. \FAMA can thus be regarded as a solving process that generates an instance of a planning model, and whose solution must be \emph{compliant} with the input plan traces. This behaviour largely differs from ML techniques, which aim to minimize an error function on the training data. Moreover, \FAMA requires far less sample data (input observations) than ML algorithms, thus alleviating the dependency on the assumption that there are enough data for learning the action models~\cite{Zhuo15}.

Like other learning approaches, \FAMA also accepts as an input an initial action model of the agent's behaviour, either complete or partially specified~\cite{ZhuoNK13,ZhuoK17}. In this case, the output model of \FAMA is compliant with both the input model and with the observations. Following this behaviour, we can define a distance measure between the two models which yields an assessment of the accuracy with which the input model explains the observations. In other words, the distance measure provides a mechanism to explicitly correct or adjust a model to the observations. By using this distance-based concept, \FAMA contributes with two novel semantic evaluation metrics that are generally more informative than counting the number of errors between two models. Moreover, this new evaluation method opens up a way towards model reconciliation~\cite{KulkarniCZVZK16}.

All in all, \FAMA is a planning-based solving scheme that outputs a \strips\ action model using an automatically built planning model from minimal input knowledge. Unlike extensive data-ML approaches, \FAMA only requires a small amount of input plan traces. Unlike most relevant action-model learning algorithms, \FAMA does not require the traces to contain any observed action executed by the agent. Ultimately, \FAMA is a model-based planning approach that adopts certain ML-like characteristics, like the ability to learn only from data that do not explicitly exhibit the agent¡s behaviour or the capacity of correcting an initially given action model.

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016generalized,segovia2016hierarchical,segovia2017generating}, this paper presents \FAMA, an innovative approach for learning \strips\ action models defined as a classical planning compilation. A first description of the compilation scheme previously appeared in our previous conference paper~\cite{aineto2018learning}. \FAMA brings the following contributions over the first version of the compilation:

\begin{itemize}
\item A unified formulation for learning and evaluating action models from observations of plan executions. In the case of minimal observability, these execution only comprise the initial and final state of the plan traces.
\item A thorough elaboration of two semantic evaluation metrics that build upon two well-known syntax-based ML metrics, {\em precision} and {\em recall}~\cite{davis2006relationship}, to evaluate the output action models with respect to observations of plan executions.
\item An exhaustive empirical evaluation over 14 domains from the International Planning Competitions (IPCs). We include an analysis of the impact of the size of the input knowledge in the accuracy and performance of \FAMA, a comparison with \ARMS and a detailed experimentation when \FAMA is executed with minimal input knowledge.
\end{itemize}

The paper is organized as follows. Section~\ref{sec:background} introduces classical planning concepts and reviews related work on learning planning action models. Section~\ref{sec:motivation} motivates our compilation-to-planning approach for learning action models. Section~\ref{sec:learning} formalizes the learning task and presents the compilation scheme, the core of \FAMA. Sections~\ref{sec:evaluation} explains the evaluation of a learned model with respect to a reference model (syntactic evaluation) and with respect to a set of plan traces (semantic evaluation). Section~\ref{sec:experiments} reports the results of the experimental evaluation and, finally, Section~\ref{sec:conclusions} discusses the strengths and weaknesses of the compilation approach and proposes several opportunities for future research.
